---
title: "RNG Slides"
author: Adam Kaplan
date: "March 28, 2022"
output: 
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "beaver"
    incremental: true
---

```{r echo=FALSE, message=FALSE}
here::i_am("rng-slides.Rmd")
library(knitr)
knit_engines$set("rcpp")
```

## Random number generation with Futures

- **Motivating example:** Monte Carlo simulations.
- To ensure replicability we would usually set a random seed using `set.seed(02135)` before running the simulations.
- Unfortunately, when we parallelize, since each process is independent of each other, it is very likely that the random state _across_ processes is no longer pseudo-random. In fact, it will likely be correlated.
- Even the naive solution of setting a different seed for each iteration does not guarantee the desired level of randomness.
- Thankfully, there are random number generation algorithms specifically designed for parallel computing. The default for `future` and most commonly used one is _L’Ecuyer CMRG algorithm_.
- Using it is as easy as in a sequential program.

## Random number generation in `future.apply`

```{r message=FALSE}
  library(future.apply)
  # Set the seed
  set.seed(02135)
  # Specify that you want to use random numbers
  out1 <- future_sapply(1:10, function(x) runif(1),
    future.seed = TRUE)
```

## Random number generation in `furrr`

```{r}
  library(furrr)
  # Set the seed
  set.seed(02135)
  # Specify that you want to use random numbers
  out2 <- future_map_dbl(1:10, ~ runif(1),
    .options = furrr_options(seed = TRUE))
  # Check that they are the same 
  all(out1 == out2)
```