---
title: "Efficient Coding in R"
author: "Adam Kaplan / Tomoya Sasaki"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "beaver"
fontsize: 10pt
---

```{r settings, echo=FALSE, message=FALSE}
here::i_am("PML Presentation/presentation.Rmd")
library(tidyverse)
knitr::knit_engines$set("rcpp")
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Efficiency in `R`

- Why is some code "fast" and other code "slow" in `R`?
- Two main culprits:
    - Bad algorithms,
    - Inefficient memory usage.
- Does not suffice to use and understand `apply()` family of functions.
- $\implies$ Need to understand basics of computational complexity and memory allocation.

## Roadmap

- Basic of computational complexity
- Memory allocation
- Memory prescriptions
- Vectorization: `apply` and `purrr`
- Parallelization: `future` and `foreach`
- Random number generator

# Computational Complexity

## Basics of Computational Complexity

- A theoretically rigorous way to measure "speed" of an algorithm as a function of input size.
- Put simply: How "fast" can $f(x_1, \dots, x_n)$ be evaluated as a function of $n$.
- Before we go any deeper, what do we mean by speed?
    - How many inputs $x_i$ does $f$ need to go over.
- Why can't we just measure time from start to end?
    - Many other factors impact the speed with which your code runs; harder to pin down issues.
- What can you do?
    1. Find more appropriate algorithms,
    2. Vectorize whenever possible (why? stay tuned),
    3. If possible, parallelize.

## Computational Complexity Examples

```{r complexity-examples1, eval=FALSE}
# Block 1: Simple for loop
for (i in 1:n) {
  foo()
}

# Block 2: Sequential for loops
for (i in 1:n) {
  foo()
}
for (i in 1:n) {
  bar()
}
```

---

```{r complexity-examples2, eval=FALSE}
# Block 3: Nested for-loops
for (i in 1:n) {
  foo()
  for (i in 1:n) {
    bar()
  }
}

# Block 4: Simple bootstrap
for (i in 1:b) {
  # Hint: mle with BFGS takes O(n^2)
  estimate <- mle(X_1, ..., X_n, method = BFGS)
}
```

# Memory Allocation in `R`

## Memory Allocation in `R`

- Now we are algorithm experts and will measure the complexity of every piece of code.
- Is this enough? Unfortunately no.
- Even if the algorithms are efficient, we can lose real-life performance due to inefficient memory usage.
- Will go over basics of memory (in `R`), data types, and vectorization to motivate efficiency prescriptions.

## What is memory?

- Computer consists of three main parts: CPU core(s) (different circuits), Memory (RAM, Hard drive), GPU (can ignore for this session)
- CPU is wicked fast.
- Many different types of memory: Volatile (RAM, Caches), Non-volatile (Hard drive, USB, CDs, ...).
- Every R session lives on RAM memory.
- RAM is actually very very fast for memory standards, but painstakingly slow compared to CPUs, particularly saving new things to it (writes).
- To write efficient code, we **need to avoid unnecessary memory writes!**

## Memory Usage in `R`

- What happens when I delete an object with `rm()`?
    - You tell `R` that this reference to an object is no longer needed.
    - If there are no more, it will be "garbage collected" _automatically_.
- **Big culprit:** Copying of objects.
    - `R` uses a "copy if modified" framework so if `y <- x`, and `y[1] <- 0`, `y` will no longer point to `x`, but will allocate a new chunk of memory and point to that.
    - $\implies$ Avoid implicit and explicit copying of objects whenever possible.
    - $\implies$ "Vectorizing" is mostly faster because it avoids this implicit copying.

## Implicit copying of objects

```{r implicit-copying, eval = FALSE}
# Example from: http://adv-r.had.co.nz/memory.html
library(pryr)
x <- data.frame(matrix(runif(100 * 1e4), ncol = 100))
medians <- vapply(x, median, numeric(1))

for (i in seq_along(medians)) {
  x[, i] <- x[, i] - medians[i]
  print(c(address(x), refs(x)))
}
```

## Explicit copying objects

- Define the size and type of object to avoid copying object.
- Example: creating a vector of ones.

---

```{r explicit-copying}
f1 <- function() { # adding ones one by one
  x <- 1.0
  for (i in 2:10000) {
    x <- c(x, 1.0)
  }
  return(x)
}
f2 <- function() { # define length + type of vector
  x <- numeric(10000) # can be character(), interger() etc
  for (i in 1:10000) {
    x[i] <- 1.0
  }
  return(x)
}
f3 <- function() { # vectorization
  x <- rep(1.0, 10000)
  return(x)
}
```

---

```{r explicit-copying-benchmark}
bench::mark( # sum two columns for each row
  f1(), f2(), f3()
) %>% select(expression, median)
```

## Functions in `R`

- From last time, arguments are not copied until modified, yay!
- What if we return a function in a function?
- **Turns out:** Functions save their surrounding environment!
- So what? Imagine you create a large variable (`x <- 1:1e+50`) before you create your function, then your function will also copy the large variable into memory.
- Now on to more applied examples of why data types matter and how to vectorize in R.

# Memory Prescriptions

## Memory Prescription 1: Data Types Matter {.build}

- List vs. Matrix vs. Dataframe (or Tibble)
    - Matrix algebra is heavily optimized!
    - Neither lists nor matrices have memory overhead.
    - Dataframes have many specialized functions such as grouping that are much faster than DIY approaches on lists or matrices.

---

```{r data-types}
x <- runif(10000)
y <- rnorm(10000)
DF <- data.frame(x = x, y = y) # dataframe
TIB <- tibble::as_tibble(DF) # tidyverse tibble
MAT <- cbind(x, y) # matrix
```

---

```{r data-types-benchmark}
bench::mark( # sum two columns for each row
  apply(DF, 2, sum),
  apply(TIB, 2, sum),
  apply(MAT, 2, sum),
  colSums(MAT) # Implemented in C
) %>% select(expression, median)
```

## Memory Prescription 2: Vectorize

- "Vectorize" your codes as much as you can.
- Most `R` functions allow you to vectorize by default.

# Vectorization: `apply` and `purrr`

## Vectorization: `apply` and `purrr`

- `apply`, `lapply`, `sapply`, ... come built-in with `R` and allow you to apply _any_ function in a vectorized way to a matrix/dataframe, list, or vector.
- `purrr`'s `map`, `map_dbl`, `map_dfc`, ... are the `tidyverse` equivalents and extensions to the `apply()` family.

---

```{r vectorization-comparison1, eval = FALSE}
library(tidyverse)
d <- list(
  data.frame(quant = c("danny", "insong", "teppei")),
  data.frame(schools = c("MIT", "Harvard"))
)
# We want a list as an output
lapply(d, function(x) nrow(x))

# default map returns a list
map(d, function(x) nrow(x))

# The tidy version
d %>% map(~ nrow(.x))
```

---

```{r vectorization-comparison2, eval = FALSE}
library(tidyverse)
v <- 1:10

# We want a list as an output
sapply(v, function(x) x * x)

# The tidy version
v %>% map_dbl(~ .x * .x)
```

## Vectorization using `purrr`

- `map_*` return different object type and will fail if it is not appropriate.
- Makes the code more predictable, and thus easier to debug.
- Of particular interest are `map_dfr` and `map_dfc`, which run a function on the input and then row/column bind the outputs together.
- Inspired popular parallelization package `furrr` which we will introduce you to shortly.

# Parallelize: `future` and `foreach`

## Why and when to parallelize

- Enable multiple computations to take place at the same time
- Useful when you have time-consuming, unordered tasks
    - Data cleaning
    - Bootstrap
    - Monte Carlo simulations
    - Any tasks with lots of loops, apply, or maps

---

## Common issues in parallel computing

- Often need two different codes for parallel and non-parallel computing
    - Increase potential bugs
- Methods for parallel computing differ across operating systems and different packages implements different methods
    - `mclapply`. `parallel`, `doParallel` etc
    - Parallelization with `mclapply` (forking) does not work in Windows
    - Your choice of package would decide how to run in parallel

---

## \texttt{future} package

- `future` provides a simple and uniform tool for async. parallel, and distributed processing in `R`
- Same coding style between sequential and parallel tasks
- Users decide how to parallelize: the code does not depend on how to run in parallel
    - Use the same code for parallel computing in different operating systems

    <!-- - In programming, "future" can be thought as a value that will eventually become available -->
    <!-- - `future` implements this idea -->

```{r idea-behind-futures, eval=FALSE}
f <- future(expr) ## evaluate in parallel
r <- resolved(f) ## check if done
v <- value(f) ## get result
```

---

## Same coding style

```{r future-example}
f3 <- function() {
  tmp <- rnorm(10000)
  for (i in 1:10000) {
    tmp <- tmp + i
  }
  sum(tmp)
}

library(future)
future_obj <- future(f3()) ## evaluate in parallel
```

---

## Choose how to parallelize

- Sequential: `plan(sequential)`
- Parallelize on local machine: `plan(multisession)`
- "Fork" your task (not fore Windows): `plan(multicore)`
- Multiple local or remote computers `plan(cluster)`

---

## Combine together

```{r}
## multisession using two cores:
## default setup uses the max # of cores
## print # of cores to use with future::availableCores()
plan(multisession, workers = 2)

future_obj <- future(f3()) ## evaluate in parallel
output2 <- value(future_obj) ## get result
output <- f3() ## non parallelized version
all.equal(output, output2)
plan() ## check current plan
```

---

## User friendly functions

- Add `future_` in front of the function
- Evaluate, check, get result within one function

```{r future-lapply-example, eval = FALSE}
## apply / future.apply
plan(multisession, workers = 2)
future_lapply(x, your_fun) ## the same as lapply(x, your_fun)
```

---

- `furrr` package: integrate with `purrr`

```{r furrr-example, eval = FALSE}
## purrr / furrr
library(furrr)
plan(sequential)
x %>% future_map(your_fun)
## the same as x %>% purrr::map(your_fun)
```

## foreach for parallel computing

- A parallel/distributed computing framework for the R language
    - A tool for "what to parallelize"
- Enables computation across multiple CPU cores and computers
- `foreach` runs sequentially in the absence of a parallel adapter
- `doFuture` from `future` framework provides a useful adapter for `foreach`

---

## Basic structure

```{r foreach-example1}
library(foreach)
## by default return a list
## where each element is an output from each iteration
res  <- foreach (i = 1:3) %do% {
  i + 1
}
res
```

---

## Basic structure 2

```{r foreach-example2}
## you can iterate over two variables with the same length
res <- foreach(i = 1:3, j = 11:13) %do% {
  i + j
}
res
```

---

## Basic structure 3

```{r foreach-example3}
## the output can be a vector
res <- foreach(i = 1:3, j = 11:13, .combine = "c") %do% {
  i + j
}
res

## the output can be pretty flexible
res <- foreach(i = 1:3, j = 11:13, .combine = "+") %do% {
  i + j
}
res
```

---

## Parallelization with `foreach` and `future`

```{r doFuture-example1}
library(doFuture)
registerDoFuture() ## register cores
plan(multisession, workers = 2) ## 2 cores

## compute mean while trimming values less than 0.1
## %dopar% automatically detect parallel adapter
cutoff <- 5
y <- foreach(i = 1:10, .combine = "c") %dopar% {
  if (i < cutoff) return(i)
}
head(y)
```

---

## xvii and `foreach`

- Your computer usually does not have enough cores for parallel computing
    - \# of your cores = \# of processors you can run simultaneously
- Use xvii, which has more than 20 cores
- The above codes work with xvii
- Do not use too many cores

# Random number generators

## Random number generation with Futures

- **Motivating example:** Monte Carlo simulations.
- To ensure replicability we would usually set a random seed using `set.seed(02135)` before running the simulations.
- Unfortunately, when we parallelize, since each process is independent of each other, it is very likely that the random state _across_ processes is no longer pseudo-random. In fact, it will likely be correlated.
- Even the naive solution of setting a different seed for each iteration does not guarantee the desired level of randomness.
- Thankfully, there are random number generation algorithms specifically designed for parallel computing. The default for `future` and most commonly used one is _L’Ecuyer CMRG algorithm_.
- Using it is as easy as in a sequential program.

## Random number generation in `future.apply`

```{r rng-future_apply, message=FALSE}
library(future.apply)
# Set the seed
set.seed(02135)
# Specify that you want to use random numbers
out1 <- future_sapply(1:10, function(x) runif(1),
  future.seed = TRUE
)
```

## Random number generation in `furrr`

```{r rng-furrr}
library(furrr)
# Set the seed
set.seed(02135)
# Specify that you want to use random numbers
out2 <- future_map_dbl(1:10, ~ runif(1),
  .options = furrr_options(seed = TRUE)
)
# Check that they are the same
all(out1 == out2)
```

## Random number generation in `foreach`

```{r, eval = FALSE}
library(foreach)
library(doRNG) ## random number generation in foreach
registerDoFuture() ## register cores
plan(multisession, workers = 2) ## multisession with 2 cores

## use %dorng% instead of %dopar%
y <- foreach(i = 1:10, .combine = "c") %dorng% {
  set.seed(i)
  rnorm(n = 1)
}
head(y)
```



# Conclusion

## Key Takeaways

- Choose your data types appropriately.
- Use canned functions whenever possible and otherwise try to avoid nested for loops if possible.
- Vectorize whenever you can.
- Be on the lookout for explicit or implicit copying of objects.

# Thank you for listening!
